% Markuss Saule - one-page ATS resume
\documentclass[letterpaper,10pt]{article}

\usepackage[letterpaper,top=0.42in,bottom=0.40in,left=0.50in,right=0.50in]{geometry}
\usepackage[hidelinks]{hyperref}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{xcolor}
\input{glyphtounicode}

\pdfgentounicode=1
\pagestyle{empty}
\urlstyle{same}
\raggedright
\setlength{\parindent}{0pt}
\setlength{\parskip}{0pt}
\setlength{\tabcolsep}{0pt}
\definecolor{rulegray}{RGB}{92,92,92}
\definecolor{meta}{RGB}{55,55,55}

\titleformat{\section}
  {\fontsize{10.5}{11.5}\selectfont\bfseries\scshape}
  {}{0pt}{}
  [\vspace{1.4pt}\color{rulegray}\titlerule]
\titlespacing{\section}{0pt}{10.5pt}{5.5pt}

\newcommand{\role}[4]{%
  \vspace{4pt}
  \begin{tabular*}{\textwidth}{@{}l@{\extracolsep{\fill}}r@{}}
    \textbf{#1} & \textcolor{meta}{#2} \\
    \textit{#3} & \textit{\textcolor{meta}{#4}} \\
  \end{tabular*}\vspace{-2pt}
}

\newcommand{\bulletlist}{%
  \begin{itemize}[
    leftmargin=0.17in,
    label={\raisebox{0.15ex}{\small -}},
    itemsep=2.1pt,
    topsep=2pt,
    parsep=0pt,
    partopsep=0pt
  ]
}
\newcommand{\bulletlistend}{\end{itemize}\vspace{3.5pt}}
\newcommand{\rbullet}[1]{\item \fontsize{9.6}{12.1}\selectfont #1}
\newcommand{\compactline}[1]{{\fontsize{9.65}{12.2}\selectfont #1}}

\begin{document}

\begin{center}
  {\fontsize{22}{24}\selectfont\bfseries MARKUSS SAULE}\\[2.5pt]
  {\fontsize{10}{11}\selectfont Business Intelligence Engineering $\mid$ Product Analytics $\mid$ Applied Data Science}\\[2.4pt]
  {\fontsize{9.3}{10.5}\selectfont
    (507) 272-7915 $\mid$
    \href{mailto:markusstomas@gmail.com}{markusstomas@gmail.com} $\mid$
    \href{https://linkedin.com/in/markuss-saule}{linkedin.com/in/markuss-saule} $\mid$
    \href{https://msaule.github.io}{msaule.github.io}}
\end{center}
\vspace{-4pt}

\section{Summary}
\compactline{Analytics and data-product builder with a 4.0 GPA and enterprise experience spanning digital measurement, statistical monitoring, applied ML, and decision systems. Owns work from instrumentation and data quality through validation, deployment, and executive-ready outcomes.}

\section{Experience}
\role{Volvo Group}{May 2026 - Present}{Business Intelligence \& Analytics Intern}{Greensboro, NC}
\bulletlist
\rbullet{Led Adobe CJA migration for digital journeys serving 3M+ monthly users across funnels and vehicle configurators.}
\rbullet{Eliminated 2 martech dependencies through tool rationalization, reducing annual software run-rate by \$50K.}
\rbullet{Built statistical monitoring for traffic, sentiment, and funnels; detected a high-intent submission drop from 9\% to 5\%.}
\bulletlistend

\role{BYU-Idaho Online Employment \& Scheduling}{Oct. 2024 - May 2026}{Data \& Campaign Analytics Coordinator}{Rexburg, ID}
\bulletlist
\rbullet{Optimized a \$4K Meta recruiting campaign through segmentation, generating 2,250 qualified applicants.}
\rbullet{Built KPI dashboards that increased hiring engagement 320\% across student and faculty audiences.}
\rbullet{Designed an AI-assisted screening workflow that standardized evaluation for 1,000+ global applicants.}
\bulletlistend

\role{Fresno State University Sports}{Jun. 2025 - Sep. 2025}{Digital Data \& Insights Coordinator}{Fresno, CA}
\bulletlist
\rbullet{Built a scoring model that ranked 350+ sponsor prospects and surfaced \$370K in sponsorship value.}
\rbullet{Automated KPI dashboards, cutting weekly reporting time 70\% and improving leadership decision speed.}
\rbullet{Analyzed 4+ content channels to guide audience, sponsorship, and campaign strategy.}
\bulletlistend

\role{Ministry of Health of Latvia}{Jun. 2021 - Oct. 2021}{Marketing \& Analytics Intern}{Riga, Latvia}
\bulletlist
\rbullet{Analyzed vaccination trends for a national COVID-19 campaign reaching 175,000 people across Latvia.}
\rbullet{Coordinated data and messaging across 3 departments to improve campaign targeting and clarity.}
\rbullet{Measured outreach performance to refine audience segments, content, and channel allocation.}
\bulletlistend

\section{Selected Projects}
\bulletlist
\rbullet{Built \textbf{Earnings Quality Autopsy}, a full-stack SEC/XBRL research system with 9 modules, 14 API routes, and 187 tests.}
\rbullet{Published \textbf{Ravel} on CRAN and built \textbf{falsifyr}, a CRAN-targeted R package for adversarial testing of statistical claims.}
\rbullet{Built \textbf{Radiology Value Pipeline}, governed clinical AI with audit trails, Power BI marts, and 0.9828 MRI AUC.}
\rbullet{Built \textbf{OWL}, a local-first AI discovery daemon with a SQLite knowledge graph, MCP server, desktop app, and 98 tests.}
\rbullet{Shipped \textbf{Roblox brand worlds} tied to 6.5B+ visits and personal recognition from CEO David Baszucki.}
\bulletlistend

\section{Education}
\role{Brigham Young University-Idaho}{Expected Apr. 2027}{B.S. Business Analytics; Minors in Statistics and Data Science; GPA: 4.0}{Rexburg, ID}

\section{Technical Skills}
\compactline{\textbf{Programming \& Data:} Python, SQL, R, pandas, scikit-learn, XGBoost, SHAP, PostgreSQL, MySQL, SQLite, FastAPI, ETL, data modeling}\\[1.2pt]
\compactline{\textbf{Analytics \& BI:} Power BI, DAX, Power Query, Adobe CJA, forecasting, simulation, experimentation, metric governance, statistical modeling}\\[1.2pt]
\compactline{\textbf{Languages:} Latvian (native), English (C2), Russian (C1) $\mid$ \textbf{Code:} \href{https://github.com/msaule}{github.com/msaule}}

\end{document}
